---
title: "Introduction to G-Computation"
author: "Nanum Jeon"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 6
)

# Load required libraries
library(dplyr)
library(knitr)
library(kableExtra)
```

# Introduction

This tutorial demonstrates g-methods in a simplified static (cross-sectional) setting, based on concepts from [this paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6074945/) by Naimi, A. I., Cole, S. R., and Kennedy, E. H (2017). By removing the time-varying complexity, we can focus on the core principles of the g-formula for causal inference.

In this tutorial, we will use the **g-formula** to estimate the average causal effect of treatment compared to no treatment. We'll demonstrate both non-parametric and parametric approaches to g-computation.

## Our Settings

We have a simple cross-sectional study examining the effect of HIV treatment ($A$) on CD4 count ($Y$). We observe a single baseline confounder, HIV viral load ($Z$), measured before treatment assignment. Unlike the longitudinal setting, all variables are measured at a single time point.

Our variables are:

- $Z$: HIV viral load at baseline (0 = low, 1 = high); 1 indicates worse baseline condition
- $A$: Treatment status (0 = no treatment, 1 = treatment)
- $Y$: CD4 count outcome; higher values indicate better immune function

We want to estimate the average causal effect of treatment ($a = 1$) compared to no treatment ($a = 0$). In this setup, the true average treatment effect (ATE) is 50.

# G-Formula in Static Setting

The g-formula allows us to estimate the average potential outcome under a hypothetical treatment intervention. In the static setting, the g-formula simplifies to:

$$ E(Y^a) = \sum_z E(Y | A = a, Z = z) \cdot P(Z = z) $$

This is essentially a standardization or adjustment formula where we:
1. Compute the expected outcome for each level of the confounder $Z$
2. Weight by the marginal distribution of $Z$ in the population

Our estimand of interest is:
$$\hat{\tau} = E[Y^{a=1}] - E[Y^{a=0}]$$

## Data Setup

```{r data-setup}
# Simplified data for static setting
# Z: Baseline HIV viral load (0 = low, 1 = high)
# A: Treatment status (0 = no treatment, 1 = treatment)  
# Y: CD4 count outcome
# N: Number of subjects in each stratum

z <- c(0, 0, 1, 1)
a <- c(0, 1, 0, 1)
y <- c(100, 150, 80, 130)
n <- c(300, 200, 150, 350)

data <- data.frame(z, a, y, n)

# Display the data
kable(data, 
      caption = "Table 1: Cross-sectional Data",
      col.names = c("Z (Viral Load)", "A (Treatment)", "Y (CD4 Count)", "N (Sample Size)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Calculate marginal distribution of Z
total_n <- sum(n)
z_marginal <- data %>%
  group_by(z) %>%
  summarise(total = sum(n), prob = sum(n) / total_n, .groups = 'drop')

kable(z_marginal,
      caption = "Table 2: Marginal Distribution of Z",
      col.names = c("Z (Viral Load)", "Count", "Probability"),
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Non-parametric G-Formula

### Treated Intervention (A = 1)

```{r treated-intervention}
# E[Y^{a=1}] - always treated
# Get expected outcomes for treated group at each level of Z
treated_outcomes <- data %>%
  filter(a == 1) %>%
  left_join(z_marginal %>% select(z, prob), by = "z") %>%
  mutate(weighted_y = y * prob)

expected_treated <- sum(treated_outcomes$weighted_y)

kable(treated_outcomes %>% select(z, y, prob, weighted_y),
      caption = "Table 3: Treated Intervention Calculation",
      col.names = c("Z", "$E[Y\\mid A=1, Z]$", "$P(Z)$", "Weighted Contribution"),
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

cat("Expected outcome under treatment: E[Y^{a=1}] =", round(expected_treated, 4), "\n")
```

### Untreated Intervention (A = 0)

```{r untreated-intervention}
# E[Y^{a=0}] - never treated
# Get expected outcomes for untreated group at each level of Z
untreated_outcomes <- data %>%
  filter(a == 0) %>%
  left_join(z_marginal %>% select(z, prob), by = "z") %>%
  mutate(weighted_y = y * prob)

expected_untreated <- sum(untreated_outcomes$weighted_y)

kable(untreated_outcomes %>% select(z, y, prob, weighted_y),
      caption = "Table 4: Untreated Intervention Calculation", 
      col.names = c("Z", "$E[Y\\mid A=0, Z]$", "$P(Z)$", "Weighted Contribution"),
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

cat("Expected outcome without treatment: E[Y^{a=0}] =", round(expected_untreated, 4), "\n")
```

### Results Summary

```{r nonparam-results}
# Calculate the average treatment effect
ate_nonparam <- expected_treated - expected_untreated

# Create results summary
results_nonparam <- data.frame(
  Intervention = c("Treated (A=1)", "Untreated (A=0)", "Average Treatment Effect"),
  Expected_Outcome = c(round(expected_treated, 4), round(expected_untreated, 4), round(ate_nonparam, 4))
)

kable(results_nonparam,
      caption = "Table 5: Non-parametric G-Formula Results",
      col.names = c("Intervention", "Expected Outcome")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

cat("Non-parametric Average Treatment Effect:", round(ate_nonparam, 4), "\n")
```

This approach represents the **non-parametric maximum likelihood estimation** of the g-formula.

# Parametric G-Formula

Now we'll demonstrate the parametric approach using regression models. This is more practical when dealing with multiple confounders or continuous variables.

## Step 1: Fit Models

```{r fit-models}
# Fit models for each component
# Model for confounder Z (baseline characteristic)
# Since Z is a baseline variable, we model its marginal distribution
z_model_data <- data %>%
  slice(rep(row_number(), n)) # Expand data based on sample sizes

# Model for treatment A given Z
a_model <- glm(a ~ z, family = binomial(), data = data, weights = n)

# Model for outcome Y given A and Z  
y_model <- lm(y ~ a + z, data = data, weights = n)

# Display model summaries
cat("Treatment Model (A ~ Z):\n")
summary(a_model)

cat("\nOutcome Model (Y ~ A + Z):\n") 
summary(y_model)
```

## Step 2: Generate Simulated Data

```{r simulate-data}
# Step 2: Generate simulated values for parametric g-formula
set.seed(12345)
sim_n <- 100000

# Simulate Z from its marginal distribution
z_sim <- sample(z_marginal$z, size = sim_n, replace = TRUE, prob = z_marginal$prob)

# Simulate A given Z (though we'll override this for interventions)
a_sim <- rbinom(n = sim_n, 
                size = 1,
                prob = predict(a_model, 
                              newdata = data.frame(z = z_sim),
                              type = "response"))

# Create simulated dataset
data_sim <- data.frame(z = z_sim, a = a_sim)

cat("Simulation completed with", sim_n, "observations\n")
cat("Z distribution:", table(z_sim), "\n")
cat("A distribution:", table(a_sim), "\n")
```

## Step 3: Intervention and Prediction

```{r parametric-intervention}
# Intervene to set treatment status
# Always treated (A = 1)
y_treated <- predict(y_model,
                    newdata = data_sim %>% mutate(a = 1),
                    type = "response")

# Never treated (A = 0)  
y_untreated <- predict(y_model,
                      newdata = data_sim %>% mutate(a = 0),
                      type = "response")

# Calculate average potential outcomes
expected_treated_param <- mean(y_treated)
expected_untreated_param <- mean(y_untreated)
ate_param <- expected_treated_param - expected_untreated_param

# Results summary
results_param <- data.frame(
  Intervention = c("Treated (A=1)", "Untreated (A=0)", "Average Treatment Effect"),
  Expected_Outcome = c(round(expected_treated_param, 4), 
                      round(expected_untreated_param, 4), 
                      round(ate_param, 4))
)

kable(results_param,
      caption = "Table 6: Parametric G-Formula Results",
      col.names = c("Intervention", "Expected Outcome")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

cat("Parametric Average Treatment Effect:", round(ate_param, 4), "\n")
```

# Comparison with Simple Methods

Let's compare our g-formula results with simpler approaches that don't properly adjust for confounding.

```{r comparison-methods}
# Crude (unadjusted) comparison
crude_treated <- data %>% filter(a == 1) %>% summarise(mean = weighted.mean(y, n)) %>% pull()
crude_untreated <- data %>% filter(a == 0) %>% summarise(mean = weighted.mean(y, n)) %>% pull()
crude_effect <- crude_treated - crude_untreated

# Summary comparison
comparison <- data.frame(
  Method = c("Crude (Unadjusted)", "Non-parametric G-Formula", "Parametric G-Formula"),
  ATE_Estimate = c(round(crude_effect, 4), 
                   round(ate_nonparam, 4), round(ate_param, 4)),
  Notes = c("Ignores confounding", "Direct standardization", "Model-based")
)

kable(comparison,
      caption = "Table 7: Comparison of Different Methods",
      col.names = c("Method", "ATE Estimate", "Notes")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```


# Summary

In this simplified static setting, we've demonstrated the core concepts of the g-formula:

1. **Non-parametric approach**: Direct standardization using the empirical distribution of confounders
2. **Parametric approach**: Model-based estimation allowing for more complex relationships

Both involve computing \( E[Y \mid A = a, Z = z] \) within levels of \( Z \) and then weighting by the marginal distribution \( P(Z = z) \). 


## Extensions

This static framework naturally extends to:

- **Multiple confounders**: Simply expand the standardization
- **Continuous confounders**: Use integration instead of summation
- **Machine learning**: Replace simple regression with flexible algorithms
- **Longitudinal settings**: Add time-varying confounders and treatments

## References

Naimi, A. I., Cole, S. R., & Kennedy, E. H. (2017). An introduction to g methods. *International Journal of Epidemiology*, 46(2), 756-762.